---
title: "Local restriction tiers in GM"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse) ; library(reactable) ; library(lemon) ; library(ggtext) ; library(ggrepel)

scores <- read_csv("data/covid19_combined_risk_scores.csv") %>% 
  arrange(desc(score)) %>% 
  mutate(rank = rank(desc(score), ties.method = "min"),
         tier = factor(tier, levels = c("Tier 3", "Tier 2", "Tier 1"), ordered = T))
```

#### Overview

This short report compares Greater Manchester's local authorities by the indicators used by Central Government to determine [local restriction tiers](https://www.gov.uk/government/news/tiering-decisions-to-be-made-through-national-command-structure) in England. The data shown cover the week ending 25 November 2020. 

#### Data and method

Local authorities have been allocated to local restriction tiers based on analysis of the following COVID-19 indicators:

|Indicator |Source |
|:--- |:--- |
|Case rate (all age groups) | [Public Health England](https://coronavirus.data.gov.uk/details/download) |
|Case rate (over-60s) | [Public Health England](https://coronavirus.data.gov.uk/details/download) |
|Case growth rate | [Public Health England](https://coronavirus.data.gov.uk/details/download) |
|Positivity rate | [NHS Test & Trace](https://www.gov.uk/government/collections/nhs-test-and-trace-statistics-england-weekly-reports) |
|Hospital admissions | [NHS](https://www.england.nhs.uk/statistics/statistical-work-areas/covid-19-hospital-activity)  |
|Hospital occupancy | [NHS](https://www.england.nhs.uk/statistics/statistical-work-areas/covid-19-hospital-activity) |

The method used by the [Joint Biosecurity Centre](https://www.gov.uk/government/groups/joint-biosecurity-centre) to allocate local authorities to tiers is not available so the following analysis adapts the [method](https://twitter.com/jburnmurdoch/status/1332233679704383488) used by [John Burn-Murdoch](https://twitter.com/jburnmurdoch) of the Finanical Times. It combines all of the indicators used to determine the local restricion tiers into a normalised risk score.

- Case, hospital admission and occupancy rates are calculated by dividing counts by estimates of the [mid-2019 population](https://www.nomisweb.co.uk/datasets/pestsyoala) multiplied by 100,000. 
- Positivity rates were calulated by dividing the number of positive confirmed cases during the latest week by the number of Pillar 1 and 2 tests. 
- Hospital occupancy and admission numbers were apportioned to local authority areas using modelled estimates of MSOA [catchment populations](https://app.powerbi.com/view?r=eyJrIjoiODZmNGQ0YzItZDAwZi00MzFiLWE4NzAtMzVmNTUwMThmMTVlIiwidCI6ImVlNGUxNDk5LTRhMzUtNGIyZS1hZDQ3LTVmM2NmOWRlODY2NiIsImMiOjh9) calculated by Public Health England for each hospital trust.

Lastly, a combined COVID-19 risk score was calculated by standardising each indicator to a z-score, averaging the scores in each local authority, and rescaling the result from 0 to 100.

#### Results

The table below shows the 7 indicators used to apportion local restriction tiers for local authorities in Greater Manchester during the week ending 25 November. The normalised score, which is scaled between 1 and 100, is shown alongside its rank relative to all other local authorities in England.

```{r table}
filter(scores, area_name %in% c("Bolton", "Bury", "Manchester", "Oldham", "Rochdale", "Salford", "Stockport", "Tameside", "Trafford", "Wigan")) %>% 
  select(-c(area_code, tier)) %>% 
  reactable(defaultPageSize = 10,
            highlight = TRUE, 
            height = "auto",
            compact = TRUE,
            borderless = FALSE,
            wrap = FALSE,
            resizable = TRUE,
            defaultSorted = "score",
            defaultSortOrder = "desc",
            defaultColDef = colDef(align = "left"),
            columns = list(
              area_name = colDef(name = "Local authority", sortable = FALSE, align = "left"),
              case_rate = colDef(name = "Case rate (all age groups)", align = "left"),
              case_rate_over_60s = colDef(name = "Case rate (over 60s)", align = "left"),
              growth_rate = colDef(name = "Case growth rate", align = "left"),
              positivity_rate = colDef(name = "Positivity rate", align = "left"),
              estimated_admissions_rate = colDef(name = "Admissions rate", align = "left"),
              estimated_occupancy_rate = colDef(name = "Occupancy rate", align = "left"),
              score = colDef(name = "Score", format = colFormat(digits = 1), align = "left"),
              rank = colDef(name = "Rank", align = "left"))
  )
```

The chart below shows the normalised combination of case rates for all groups and over 60s, case rate growth, positivity rate, hospital admissions and bed occupancy for each local authority in England by rank order. Each local authority is colour coded by the local tier restrictions that they currently sit within.

```{r scores, fig.width = 10, dev = 'svg'}
ggplot(scores, aes(rank, score, fill = factor(tier))) +
  geom_point(size = 3, pch = 21, colour = "#FFFFFF", alpha = 0.8, show.legend = FALSE) +
  geom_point(data = filter(scores, area_name %in% c("Bolton", "Bury", "Manchester", "Oldham", "Rochdale", "Salford", "Stockport", "Tameside", "Trafford", "Wigan")),
             aes(rank, score, fill = factor(tier)), size = 3, pch = 21, colour = "#000000", alpha = 0.8, show.legend = FALSE) +
  geom_text_repel(data = filter(scores, area_name %in% c("Bolton", "Bury", "Manchester", "Oldham", "Rochdale", "Salford", "Stockport", "Tameside", "Trafford", "Wigan")),
                  aes(rank, score, label = area_name), size = 3, point.padding = 0.5, hjust = 0, colour = "#000000", show.legend = FALSE) +
  scale_fill_manual(values = c("#D93621", "#EF6F2D", "#F5B63F")) +
  scale_y_continuous(breaks = seq(0,100,10)) +
  labs(x = "Rank of combined COVID-19 risk score",
       y = NULL,
       title = "Combined COVID-19 risk score",
       subtitle = paste0("<span style = 'color:#757575;'>Restriction tiers by local authority, week ending 25 November</span>"),
       caption = "Normalised score of local restriction tier indicators\nData: PHE; NHS; DHSC; ONS\nCredits: @jburnmurdoch; @carlbaker; @VictimOfMaths\nPlot by @traffordDataLab") +
  facet_rep_grid(. ~ tier, scales = "free", space = "free") +
  theme_minimal() +
  theme(plot.margin = unit(rep(0.5, 4), "cm"),
        panel.spacing = unit(0, "lines"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_markdown(size = 12, margin = margin(b = 20)),
        plot.caption = element_text(colour = "grey60", margin = margin(t = 20, b = -10)),
        strip.text.x = element_text(size = 12, face = "bold", hjust = 0),
        axis.title.x = element_text(size = 10),
        axis.text.x = element_blank(),
        axis.text.y = element_text(face = "bold"),
        axis.line.y = element_line(colour = "#bdbdbd")) +
  coord_cartesian(clip = "off") +
  xlim(0,320)
```
